Data Specialization for a Staged Lambda Calculus

### 2015-02-28 paper/abstract.tex
Editing pass on abstract (to tighten a bit)

Staged programming languages assign a stage to each expression and
evaluate each expression in its assigned stage.  For example, in a
two-staged program, some expressions are evaluated in the first stage
and others in the second stage.  Programming languages, evaluation
strategies, and type systems for staged programs where stages roughly
correspond to the arrival order of inputs have been studied
extensively in the areas of partial evaluation and meta programming.

In this paper, we consider staged programs written in a typed lambda
calculus with a $\fut$ modality denoting computation in a later
stages and present techniques for splitting them into multi-pass
programs. Such multi-pass programs evaluate in a sequence of
evaluations of unstaged, conventional programs each of which
corresponds to a pass.  Conceived in the 80's by J{\o}rring and
Scherlis, this {\em pass-separation} technique has been studied, but
existing techniques are either manual or applicable only to simple programs
that lack important features such as recursion and first-class
functions.

% We make no restrictions on the kinds of programs that can
% be written.  We also present an implementation.
% %

Since we build on a lambda calculus, our techniques enable the
programmer to write staged programs by using powerful programming
abstractions in a composable and modular fashion. Our splitting
transformation maps such programs into multi-pass programs that can
run significantly more efficiently by assigning frequent computations
to earlier passes so that they can be reused (cheaply) later.  Our
experiments based on a prototype implementation show that our
splitting algorithm is able perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.

>>>>

Staged programming languages assign a stage to each expression and
evaluate each expression in its assigned stage.
%For example, in a
%two-staged program, some expressions are evaluated in the first stage
%and others in the second stage.
%Programming languages, evaluation
%strategies, and type systems for staged programs where stages roughly
%correspond to the arrival order of inputs have been studied
%extensively in the areas of partial evaluation and meta programming.
While staged programming has been studied extensively in the areas of
partial evaluation and meta programming, in this paper we present an
algorithm for statically {\em splitting} programs witten in a
two-stage typed lambda calculus (with a $\fut$ modality denoting
computation in a later stage) into multi-pass programs that are
structured as two dependent, unstaged programs and are evaluated using
two distinct execution passes.  While splitting, also called pass
separation or data specialization, has been studied in the past, prior
techniques were either manual or applicable only to simple programs
that lack important features such as recursion and first-class
functions.

Since our work builds on a lambda calculus, our splitting algorithm
enables the programmer to write staged programs by using powerful
programming abstractions in a composable and modular fashion. Our
splitting transformation maps such programs into multi-pass programs
that can run significantly more efficiently by assigning frequent
computations to earlier passes so that they can be reused (cheaply)
later.  Our experiments based on a prototype implementation show that
our splitting algorithm is able perform highly non-trivial algorithmic
transformations, sometimes automatically improving efficiency
asymptotically by nearly a linear factor.  Since splitting
transformation takes place at compile time, these benefits are
achieved without significant run-time penalty.
